Galois Characterization of Function Fields of \\ Transcendence Degree $1$ over Number Fields

### 2017-07-26 Master Thesis/introduction.tex
Fix typos
Natural numbers start with 1
Finish second part of Lemma 2.2
Small fixes

Finite fields have an abelian absolute Galois group. In fact, they all have the same absolute Galois group $\hat{\mathbb{Z}}$, therefore not much information can be retrieved from these groups. The study of these types of questions belong in the area of \textit{anabelian geometry}. This term refers to the fact that the less abelian $\pi_1^\text{ét}(X)$ is, the more information it carries about $X$.

@@

This has been proved by F. Pop (see \cite{Po90}, \cite{Po94}, \cite{Po95}) using model theory. In this thesis, we want to follow M. Spiess \cite{Sp96} to prove the statement for finitely generated fields with transcendence degree $1$ over $\mathbb{Q}$, using a purely arithmetical approach and a direct generalization of the original proof of Theorem \ref{thm:neukirch-uchida}:

@@

In the first chapter, we want to recap some results from valuation theory. In the second part, we will prove some lemmata in order to be able to characterize henselian fields over $K$ of valuations of rational rank $2$ in terms of their Galois cohomology groups of dimension $3$.

@@

$\begin{array}{ll}
\mathbb{N} & \text{natural numbers} \\
\mathbb{Z} & \text{integers}\\
\mathbb{Q} & \text{rational numbers}\\
\mathbb{R} & \text{real numbers}\\
\mathbb{C} & \text{complex numbers}\\
\mathbb{F}_q & \text{finite field with $q$ elements}\\
\mathbb{Z}_p & \text{$p$-adic integers}\\
\hat{\mathbb{Z}} & \text{profinite completion of $\mathbb{Z}$}\\
\\
\overline{K} & \text{algebraic closure of $K$}\\
K^\text{s} & \text{separable closure of $K$}\\
K^\text{t} & \text{absolute inertia field of $K$}\\
K_v & \text{henselization of $K$ with respect to $v$}\\
G_K & \text{absolute Galois group of $K$}\\
\operatorname{Gal}(L/K) & \text{Galois group of $L/K$}\\
\operatorname{char}(K) & \text{characteristic of $K$}\\
\operatorname{trdeg}(L/K) & \text{transcendence degree of $L/K$}\\
\mu_n & \text{$n$-th roots of unity in a separable closure}
\end{array}$

>>>>

Finite fields have an abelian absolute Galois group. In fact, they all have the same absolute Galois group $\widehat{\mathbb{Z}}$, therefore not much information can be retrieved from these groups. The study of these types of questions belong to the area of \textit{anabelian geometry}. This term refers to the fact that the less abelian $\pi_1^\text{ét}(X)$ is, the more information it carries about $X$.

@@

This has been proven by F. Pop (see \cite{Po90}, \cite{Po94}, \cite{Po95}) using model theory. In this thesis, we want to follow M. Spiess \cite{Sp96} to prove the statement for finitely generated fields with transcendence degree $1$ over $\mathbb{Q}$, using a purely arithmetical approach and a direct generalization of the original proof of Theorem \ref{thm:neukirch-uchida}:

@@

In the first chapter, we want to recap some results from valuation theory. In the second part, we will prove some lemmata in order to characterize henselian fields over $K$ of valuations of rational rank $2$ in terms of their Galois cohomology groups of dimension $3$.

@@

$\begin{array}{ll}
\mathbb{N} & \text{natural numbers starting from $1$} \\
\mathbb{Z} & \text{integers}\\
\mathbb{Q} & \text{rational numbers}\\
\mathbb{R} & \text{real numbers}\\
\mathbb{C} & \text{complex numbers}\\
\mathbb{F}_q & \text{finite field with $q$ elements}\\
\mathbb{Z}_p & \text{$p$-adic integers}\\
\widehat{\mathbb{Z}} & \text{profinite completion of $\mathbb{Z}$}\\
\\
\overline{K} & \text{algebraic closure of $K$}\\
K^\text{s} & \text{separable closure of $K$}\\
K^\text{t} & \text{absolute inertia field of $K$}\\
K_v & \text{henselization of $K$ with respect to $v$}\\
G_K & \text{absolute Galois group of $K$}\\
\operatorname{Gal}(L/K) & \text{Galois group of $L/K$}\\
\operatorname{char}(K) & \text{characteristic of $K$}\\
\operatorname{trdeg}(L/K) & \text{transcendence degree of $L/K$}\\
\mu_n & \text{$n$-th roots of unity in a separable closure}
\end{array}$

### 2017-07-26 Master Thesis/synopsis.tex
Small fixes

In this section we shall recall facts from valuation theory.

@@

\begin{proof}
See \cite{EP05} Section 2.3, Corollary 4.1.4, \cite{ZS60} VI, §3 Theorem 3.
\end{proof}

@@

\begin{proof}
See \cite{EP05} Lemma 5.2.1, Lemma 5.2.2, Theorem 5.2.5.
\end{proof}

@@

\begin{lemma}\label{2.2}
Let $K$ be henselian with respect to two different valuations $w_1,w_2\in |\operatorname{Spv}(K)|$. Then $K$ is algebraically closed or $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K),\ v(w_1)=v(w_2)$ and $k(v(w_1))$ is separably closed. In this case, we have $\operatorname{cd}(G_K)\leq 2$.
\end{lemma}

@@

\begin{proof}
(i) and (ii) follow from \cite{HK94} Lemma 2.4.3, and (iii) is a consequence of (ii) and the fact that henselizations commute with filtrated direct limits, e.g. see \cite{EGA4} Proposition 18.6.14.
\end{proof}

@@

Let $\mathcal{P}(F)$ denote the group of principal ideals of $F$, $\mathcal{I}(F)$ the ideal group of $F$, and $\operatorname{Cl}(F) = \mathcal{I}(F)/\mathcal{P}(F)$ the ideal class group of $F$. We have the following exact sequence:
\[ \begin{tikzcd}
1 \ar[r]& \mathcal{O}_{F_1}^\times /\mathcal{O}_F^\times\ar[r] & F_1^\times / F^\times\ar[r] & \mathcal{P}(F_1)/\mathcal{P}(F)\ar[r] & 1
\end{tikzcd} \]
Again by Lemma \ref{5.2} (ii), it suffices to show that $\mathcal{P}(F_1)/\mathcal{P}(F)$ is free-by-finite. 
Consider the following diagram with exact rows and columns:
\[ \begin{tikzcd}
& 0 \ar[d] & 0 \ar[d] & \ker(f) \ar[d]& \\
0\ar[r] & \mathcal{P}(F) \ar[d] \ar[r]& \mathcal{I}(F) \ar[r]\ar[d]& \operatorname{Cl}(F)\ar[r]\ar[d, "f"] & 0\\
0 \ar[r]& \mathcal{P}(F_1)\ar[d] \ar[r]& \mathcal{I}(F_1) \ar[d]\ar[r]& \operatorname{Cl}(F_1)\ar[d] \ar[r]& 0 \\
& \mathcal{P}(F_1)/\mathcal{P}(F) & \mathcal{I}(F_1)/\mathcal{I}(F) & \operatorname{coker}(f) &
\end{tikzcd} \]
Using the snake lemma, we get the following exact sequence:
\[
0 \longrightarrow \ker(f)\longrightarrow \mathcal{P}(F_1)/\mathcal{P}(F)\longrightarrow \mathcal{I}(F_1)/\mathcal{I}(F)
\longrightarrow \operatorname{coker}(f)\longrightarrow 0 \]
Since the class numbers of number fields are finite, according to Lemma \ref{5.2} (i), it suffices to show that $\mathcal{I}(F_1)/\mathcal{I}(F)$ is free-by-finite. This follows from the fact that there are only finitely many primes of $F$ which are ramified in $F_1$.\qedhere
\end{enumerate}
\end{proof}

>>>>

In this section we shall recall some facts from valuation theory.

@@

\begin{proof}
See \cite{EP05} Section 2.3 and Corollary 4.1.4, \cite{ZS60} VI, §3 Theorem 3.
\end{proof}

@@

\begin{proof}
See \cite{EP05} Lemma 5.2.1, Lemma 5.2.2, and Theorem 5.2.5.
\end{proof}

@@

\begin{lemma}\label{2.2}
Let $K$ be henselian with respect to two different valuations $w_1,w_2\in |\operatorname{Spv}(K)|$. Then $K$ is algebraically closed or $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K),\ v(w_1)=v(w_2)$ and $k(v(w_1))$ is separably closed. In this case, we have $\operatorname{cd}(K)\leq 2$.
\end{lemma}

@@

\begin{proof}
(i) and (ii) follow from \cite{HK94} Lemma 2.4.3 and the fact that singularities of two-dimensional $\mathbb{Z}$-schemes of finite type can be resolved. (iii) is a consequence of (ii) and the fact that henselizations commute with filtrated direct limits, e.g. see \cite{EGA4} Proposition 18.6.14.
\end{proof}

@@

Let $\mathcal{P}(F)$ denote the group of principal ideals of $F$, $\mathcal{I}(F)$ the ideal group of $F$, and $\operatorname{Cl}(F) = \mathcal{I}(F)/\mathcal{P}(F)$ the ideal class group of $F$. We have the following exact sequence:
\[ \begin{tikzcd}
1 \ar[r]& \mathcal{O}_{F_1}^\times /\mathcal{O}_F^\times\ar[r] & F_1^\times / F^\times\ar[r] & \mathcal{P}(F_1)/\mathcal{P}(F)\ar[r] & 1
\end{tikzcd} \]
Again by Lemma \ref{5.2} (ii), it suffices to show that $\mathcal{P}(F_1)/\mathcal{P}(F)$ is free-by-finite. 
Consider the following diagram with exact rows:
\[ \begin{tikzcd}
0\ar[r] & \mathcal{P}(F) \ar[d] \ar[r]& \mathcal{I}(F) \ar[r]\ar[d]& \operatorname{Cl}(F)\ar[r]\ar[d, "f"] & 0\\
0 \ar[r]& \mathcal{P}(F_1) \ar[r]& \mathcal{I}(F_1) \ar[r]& \operatorname{Cl}(F_1) \ar[r]& 0 
\end{tikzcd} \]
Using the snake lemma, we get the following exact sequence:
\[
0 \longrightarrow \ker(f)\longrightarrow \mathcal{P}(F_1)/\mathcal{P}(F)\longrightarrow \mathcal{I}(F_1)/\mathcal{I}(F)
\longrightarrow \operatorname{coker}(f)\longrightarrow 0 \]
Since the class numbers of number fields are finite, according to Lemma \ref{5.2} (i), it suffices to show that $\mathcal{I}(F_1)/\mathcal{I}(F)$ is free-by-finite. This follows from the fact that there are only finitely many primes of $F$ which are ramified in $F_1$.\qedhere
\end{enumerate}
\end{proof}
